title: "Phosphorus Traits"
author: "Jay T. Lennon, Mario Muscarella, Kali Bird"
date: "`r format(Sys.time(), '%d %B, %Y')`"
header-includes:
  - \usepackage{array}
output: pdf_document
geometry: margin=2.54cm
---

Overview: make heatmaps with our data to visualize use of different phosphorus (P) sources by bacteria from an oligotrophic lake (Little Long) and eutrophic lake (Wintergreen Lake). First panel (upper left) randomizes the the data by row and column. Second panel (upper right) randomizes so there are differences among strains (not P sources). Third panel (lower left) randomizes so there are difference among P sources (not strains). Fourth panel (lower right) is the observed data. 

# Set working directory
```{r}
rm(list=ls())
getwd()
setwd("~/GitHub/PhosTrait")
```

# Load packages
```{r}
require("png")
require("vegan")
require("pvclust")
require("gplots")
require("gridGraphics")
require("grid")
require("gplots")
require("gridExtra")
require("viridis")
```

# Load data 
```{r}
# Raw growth rate data
gr.data <- read.csv("data/grraw.csv", sep=",", header=TRUE)
str(gr.data)
colnames(gr.data)[1]<-"isolate"

# Log transformed raw data 
gr.raw <- gr.data[,2:19]
gr.raw.log <- log10(gr.raw + 0.001)

# Standardizing 1: divide each observation by sum of species growth
gr.std <- gr.data[,2:19] / (apply(gr.data[,2:19], 1, sum))

# Standarizing 2: center and scale log-transformed growth
gr.std2 <- (gr.raw.log - mean(unlist(gr.raw.log))) # / sd(unlist(gr.raw.log))

# Names
resources <- colnames(gr.raw)
strains <- gr.data[,1]
```

# Customize color palette
```{r}
jet.col <- colorRampPalette(c("#00007F", "blue", "#007FFF", "cyan", 
                                 "#7FFF7F", "yellow", "#FF7F00", "red", 
                                 "#7F0000"))

gray.col <- gray.colors(25, start = 0.9, end = 0.1, gamma = 1.5, alpha = 1)

yelred.col <- colorRampPalette(c("yellow","orange", "red"))

green.col <- colorRampPalette(c("#e5f5e0", "#c7e9c0", "#a1d99b", 
                        "#74c476", "#41ab5d", "#238b45", "#00441b"))

viridis.col <-colorRampPalette(viridis(12))

colors <- viridis.col
```

# Clustering parameters
```{r}
distance <- "euclidean" 
    # others to consider: manhattan, correlation, uncentered,
    # abscor, euclidean, bray

# Identify agglomerative method
clust <- "average"   
    # others to consider:
    # complete = furthest neighbor; compact clusters, sensitive to outliers
    # average = UPGMA; considered robust
    # ward.D = popular, but makes clusters of equal size, sensitive to outliers
    # ward.D2 = dissimilarities are squared before clustering
    # mcquitty
    # median = downweights outliers
    # centroid = fairly robust
    # single = nearest neighbor; chaining problem
```

# Create four data sets for plotting
```{r}
# Create matrix (add mean back to scale to zero)
mat <- as.matrix(gr.raw.log) + (-mean(as.matrix(gr.raw.log)))

# Quad 1 - Randomize on all
full_rand <- matrix(sample(unlist(mat)), nrow = nrow(mat))

# Quad 2 - Simulate on row means
row_rand <- matrix(NA, nrow = nrow(mat), ncol = ncol(mat))
for(i in 1:nrow(row_rand)){
  temp.mean <- mean(mat[i, ])
  temp.sd <- sd(mat[i, ])
  row_rand[i, ] <- rnorm(n = dim(mat)[2], mean = temp.mean, sd = temp.sd * 0.1)
}

# Quad 3 - Randomize on column means
col_rand <- matrix(NA, nrow = nrow(mat), ncol = ncol(mat))
for(i in 1:ncol(col_rand)){
  temp.mean <- mean(mat[, i])
  temp.sd <- sd(mat[, i])
  col_rand[, i] <- rnorm(n = dim(mat)[1], mean = temp.mean, sd = temp.sd * 0.1)
}

# Quad 4 - Observed data
no_rand <- mat
```

# Make multipanel heatmap
```{r}
# Example: http://stackoverflow.com/questions/13081310/combining-multiple-complex-plots-as-panels-in-a-single-figure

# Make a list of dataframes for each scenario
heat.list <- list(full_rand, row_rand, col_rand, no_rand)

# Function for running through list
grab_grob <- function(){
  grid.echo()
  grid.grab()
}

gl <- lapply(1:4, function(i){
  heatmap.2(as.matrix(heat.list[[i]]), distfun = function(x)  
          vegdist(x, method = distance), hclustfun = function(x)
          hclust(x, method = clust), key = FALSE, keysize = 1,
          key.title = NA, key.xlab = NA, col = colors, trace = "none", 
          density.info = "none", labRow = strains, 
          labCol = resources, symkey = FALSE, 
          colsep = 1:16, rowsep = 1:37, sepcolor = "white",
          scale = "none", cexRow = 1, cexCol = 1, margins = c(6,6),
          lmat = rbind( c(0, 3), c(2,1), c(0,4) ), lhei = c(0.25, 4, 0.25 ),                 
  )
  grab_grob()
})

png(filename="~/GitHub/PhosTrait/figures/HeatMap.png", width = 2400, 
    height = 2400, res = 96*2)

grid.newpage()
grid.arrange(grobs = gl, ncol = 2, clip = TRUE)

dev.off()
graphics.off()
```
