title: "Phosphorus Traits - Quantifying Phylogenetic Signal"
author: "Jay T. Lennon, Mario Muscarella, Kali Bird"
date: "`r format(Sys.time(), '%d %B, %Y')`"
header-includes:
  - \usepackage{array}
output: pdf_document
geometry: margin=2.54cm
---

# Set working directory
```{r}
rm(list=ls())
getwd()
setwd("~/GitHub/PhosTrait")
```

# Load packages
```{r}
tidy.opts=list(width.cutoff=60)} 
package.list <- c('ape', 'seqinr', 'muscle', 'phylolm', 'vegan', 'grid', 'png', 
      'phylobase', 'adephylo', 'geiger', 'picante', 'stats', 'RColorBrewer', 
      'caper', 'phylolm', 'pmc', 'ggplot2', 'tidyr', 'dplyr', 'phangorn', 'pander') 
for (package in package.list) {
  if (!require(package, character.only=TRUE, quietly=TRUE)) {
    install.packages(package)
    library(package, character.only=TRUE)
  }
}
```

# Load growth rate data
```{r}
# Load raw growth rate data
gr.data <- read.csv("data/grraw.csv", sep=",", header = TRUE, row.names = 1)
str(gr.data)

# Standadize growth rates across strains
p.growth.std <- gr.data / (apply(gr.data, 1, sum))

# Calculate max growth rate
umax <- as.matrix((apply(gr.data, 1, max))) 
round(apply(gr.std, 2, max), 4)
round(apply(gr.std, 2, min), 4)
```

# Calculate niche breadth (Levins 1968)
```{r}
levins <- function(p_xi = ""){
  p = 0
  for (i in p_xi){
    p = p + i^2
    }
  nb = 1 / (length(p_xi) * p)
  return(nb)
  }

# Calculate niche breadth for Each Isolate
nb <- as.matrix(levins(gr.std))

# Add Row & Column Names to Niche Breadth Matrix
rownames(nb) <- rownames(gr.data)
colnames(nb) <- c("nb")
```  

# Load Maximum Likelihood tree
```{r}
# Keep rooted tree but drop outgroup branch
ml.bootstrap.out <- read.tree("./data/ml_tree/RAxML_bipartitions.T1")
ml.bootstrap <- drop.tip(ml.bootstrap.out, "Methanosarcina")
```

# Test for phylogenetic signal: Pagel's lambda

Pagel's lambda quantifies signal via transformation that is accomplished by scaling the off-diagonal elements of the variance-covariance matrix, which describes the tree topology and branch lengths with the value “lambda”. 
When lambda equals 1, you have a tree corresponding to Browninan motion, which is equivalent to your observed tree. 
When lambda equals 0, you have removed all phylogenetic signal from your tree.
When lambda equals 0.5, there is an intermediate amount of phylogenetic signal in your tree.

```{r}
# Visualize trees with different Levels of phylogenetic signal {geiger}
ml_1.0 <- rescale(ml.bootstrap, "lambda", 1)
ml_0.5 <- rescale(ml.bootstrap, "lambda", 0.5)
ml_0.0 <- rescale(ml.bootstrap, "lambda", 0)

layout(matrix(c(1,2,3), 1, 3), width = c(1, 1, 1))
par(mar=c(1,0.5,2,0.5)+0.1) 
plot(ml_1.0, main = "lambda = 1", cex = 0.7, adj = 0.5) 
plot(ml_0.5, main = "lambda = 0.5", cex = 0.7, adj = 0.5)
plot(ml_0.0, main = "lambda = 0",  cex = 0.7, adj = 0.5)

# Generate test statistics for comparing phylogenetic signal {geiger}

fitContinuous(ml.bootstrap, nb, model = "lambda") #  
# Lambda: fits extent to which phylogeny predicts covariance among trait values for species
# AIC = -33.073199

fitContinuous(ml.bootstrap, nb, model = "BM") 
# Brownian model: correlation structure among traits proportion shared ancestry for pairs of taxa
#AIC = -35.073199

fitContinuous(ml.bootstrap, nb, model = "OU") 
# Ornstein-Uhlenbeck: fits a random walk witha central tendency with an attraction strength proportional to the parameteralpha
# AIC = 263.930359 --> Warnings!

fitContinuous(ml.bootstrap, nb, model = "EB") 
# Early-burst model
# AIC = 268.500484

fitContinuous(ml.bootstrap, nb, model = "trend") 
# Diffusion model with linear trend in rates through time 
# AIC = 272.563618

fitContinuous(ml.bootstrap, nb, model = "kappa") 
# Punctuational model of trait evolution; characterd ivergence related to number of speciation events between two species
# AIC = -24.820803

fitContinuous(ml.bootstrap, nb, model = "drift") 
# Directional drift component 
# AIC = 273.580391

fitContinuous(ml.bootstrap, nb, model = "white") 
# white noise; data from single normal distribution with no covariance structure among specie
# AIC = -37.500458

### There ares some other options for specifiying models
### https://cran.r-project.org/web/packages/geiger/geiger.pdf
### There are some issues with Pagel's lambda, too:
### https://www.carlboettiger.info/2013/10/11/is-it-time-to-retire-pagels-lambda.html
```

# Test for phylogenetic signal: Blomberg's K

Statistic that quantifies phylogenetic signal by comparing observed trait
distributions on a tree to what would evolve under Brownian (i.e., random) motion.
Blomberg’s K is calculated as the mean squared error of the tip data (i.e., trait) measured from the phylogeneticcorrected mean and the mean squared error based on the variance–covariance matrix derived from the given phylogeny under the assumption of Brownian motion.
K-value of 1 means that a trait is distributed on the tree according to null expectation of Browning motion.
A K-value > 1 means that traits are clustered with closely related species being more similar than expected by chance. 
A K-value of < 1 means that traits are overdispersed with closely related species less similar
than expected by chance.

```{r}
# First, correct for zero branch-lengths on tree
ml.bootstrap$edge.length <- ml.bootstrap$edge.length + 10^-7

# Calculate phylogenetic signal for growth on all P resources
# Create blank output matrix
p.phylosignal <- matrix(NA, 6, 18)
colnames(p.phylosignal) <- colnames(p.growth.std)
rownames(p.phylosignal) <- c("K", "PIC.var.obs", "PIC.var.mean", 
                             "PIC.var.P", "PIC.var.z", "PIC.P.BH")

# Use for loop to calculate Blomberg's K for each resource
for (i in 1:18){
  x <- as.matrix(p.growth.std[ ,i, drop = FALSE])
  out <- phylosignal(x, ml.bootstrap)
  p.phylosignal[1:5, i] <- round(t(out), 3)
}

# Use the BH correction on P-values: 
p.phylosignal[6, ] <- round(p.adjust(p.phylosignal[4, ], method = "BH"), 3)

# PIC = phylogenetically independent contrasts
# K = K statistics
# PIC.var.obs	= observed PIC variance
# PIC.var.mean = mean observed PIC variance
# PIC.var.P = p-value of observed vs. random variance of PIC
# PIC.variance z = z-score of observed vs. random variance of PICs
# PIC.var.P.BH = BH-corrected p-value

# Interpretation of results:
# K values are small (~0) sugggesting that P tends towards overdispersion
# Without BH correction, this significant (alpha = 0.05) for AEP and cAMP, with marginal effects for DNA
# After BH-correction, however, there is no significant overdispersion. 

signal.nb <- phylosignal(nb, ml.bootstrap)
# again, trend towards overdispersion, but not significant 
```
