title: "Phosphorus Traits"
author: "Jay T. Lennon, Mario Muscarella, Kali Bird"
date: "`r format(Sys.time(), '%d %B, %Y')`"
header-includes:
  - \usepackage{array}
output: pdf_document
geometry: margin=2.54cm
---

Can we use our data to classify groups of phoshporus compounds?
Do these clusters make sense based on theories developed elsewhere (Muscarella et al. 2014)?
Are P sources substituable?

```{r}
rm(list=ls())
getwd()
setwd("~/GitHub/PhosTrait")
```

# Load package
```{r}
tidy.opts=list(width.cutoff=60)  
package.list <- c('ape', 'seqinr', 'muscle', 'phylolm', 'vegan', 'grid', 'png', 
      'phylobase', 'adephylo', 'geiger', 'picante', 'stats', 'RColorBrewer', 'mclust',
      'caper', 'phylolm', 'pmc', 'ggplot2', 'tidyr', 'dplyr', 'phangorn', 'pander', 'pvclust') 
for (package in package.list) {
  if (!require(package, character.only=TRUE, quietly=TRUE)) {
    install.packages(package)
    library(package, character.only=TRUE)
  }
}
```

# Load data and standardize 
```{r}
# Raw growth rate data
gr.raw.data <- read.csv("data/grraw.csv", sep=",", header=TRUE)
str(gr.raw.data)
colnames(gr.raw.data)[1]<-"isolate"

# Data standardizing - log10 transformation
gr.raw.log <- log10(gr.raw.data[,2:19]+0.001)

# Data standardizing - divide by sum of species growth
gr.std <- gr.raw.data[,2:19] / (apply(gr.raw.data[,2:19], 1, sum))

# Choose data for clustering
data <- gr.raw.log
```

# Hierarchical clustering
Here we use aglomerative clustering, which minimizes within-cluster sums-of-squared
distances between P resources. 

Step 1: identify a distance metric. 
Common metrics are `euclidean`, `manhattan`, `correlation`, `uncentered`, and `abscor`

```{r}
# Identify distance metric
dist <- "euclidean" 
```

Step 2: identify agglomerative clustering method:

```{r}
clust <- "average"   
    # complete = furthest neighbor; compact clusters, sensitive to outliers
    # average = UPGMA; considered robust
    # ward.D = popular, but makes clusters of equal size and sensitive to outliers
    # ward.D2 = dissimilarities are squared before clustering
    # mcquitty
    # median = downweights outliers
    # centroid = fairly robust
    # single = nearest neighbor; chaining problem
```
 
Step 3: perform hierarchical cluster analysis.
Reduce nboot to 100 or 1000 when playing around. 
Run 10,000 after (time consuming)

```{r}
pv.clust <- pvclust(data, nboot = 10000, method.dist = dist, 
            method.hclust = clust)
```

Notes on pvclust function (http://stat.sys.i.kyoto-u.ac.jp/prog/pvclust/):
Assesses uncertainty in hierarchical cluster analysis. 
For each cluster, p-values are calculated via multiscale bootstrap resampling. 
P-value of a cluster is a value between 0 and 1, which indicates how strong the cluster is supported by data where higher values (e.g., 92) represent more support.

The function provides two types of P-values: AU (Approximately Unbiased) P-values and BP (Bootstrap Probability) values. 
AU P-values, which are computed by multiscale bootstrap resampling, are better approximations for unbiased P-value than BP value computed by normal bootstrap resampling. 

For a cluster with AU p-value > 0.95, the hypothesis that "the cluster does not exist" is rejected with significance level 0.05

# Plotting
```{r}
# Initiate file for figure
png(filename="~/GitHub/PhosTrait/figures/Cluster.png",
    width = 1200, height = 1200, res = 96*2)

# Make plot
plot(pv.clust, main = NA, sub = NA, cex = 1.0, cex.pv = 1, 
     col.pv = c("black", "white", "black"), print.pv = TRUE, 
     print.num = FALSE, float = 0.02, lwd = 2, xlab = NA, 
     cex.lab = 1.5, las  = 2)

# Highlights clusters based on P-values
# That is, clusters don't exist in highlighted regions
# Type "geq" clusters with p-value >= than alpha threshold
# Type "leq" clusters with p-value <= than alpha threshold

pvrect(pv.clust, alpha = 0.95, pv = "au", type = "geq", 
       border = "grey", lwd = 2.5, lty = 2) 

dev.off()
graphics.off() 

img <- readPNG("./figures/Cluster.png")
grid.raster(img)

# Diagnostic plot for standard error of p-value
print(pv.clust)
seplot(pv.clust)
# Standard errors are small
```

# Plot ranked growth on different P sources

```{r}

# Calcualte mean, SEM, and sort for plotting
p.means <- apply(gr.raw.data[,2:19], 2, mean)
sem <- function(x, ...){sd(x, na.rm = TRUE)/sqrt(length(na.omit(x)))}
p.sem <- apply(gr.raw.data[,2:19], 2, sem)
p.95.LL <- apply(gr.raw.data[,2:19], 2, FUN = function(x) t.test(x)$conf.int[1])
p.95.UL <- apply(gr.raw.data[,2:19], 2, FUN = function(x) t.test(x)$conf.int[2])
p.combo <- data.frame(p.means, p.sem, p.95.LL, p.95.UL)
p.sort <- p.combo[order(p.combo$p.means),]
p.rank.name <- rownames(p.sort)

# For plotting only, we set zero values to 0.001
gr.raw.data[gr.raw.data==0]<-0.0001

# Reorder columns (p sources) of raw data by sorted rank
gr.raw.data.rank<-gr.raw.data[p.rank.name]
j <- rank(colMeans(gr.raw.data.rank))
# gr.raw.data.rank <- data.frame(j,gr.raw.data.rank)

# Plotting

# Initiate file for figure
png(filename="~/GitHub/PhosTrait/figures/Ranked.P.png",
    width = 1200, height = 800, res = 96*2)

par(mar = c(4, 5, 1, 1))

p.plot <- plot(x = rnorm(dim(gr.raw.data.rank)[1], 1, 0.05), 
               y = gr.raw.data.rank[,1], 
               ylim = c(0.00001, 5), xlim = c(0, 18), log = "y",
               pch = 21, col = "lightgrey", bg = "lightgrey", 
               lwd = 2, cex = 1.7, 
               yaxt = "n", xaxt = "n", cex.lab = 2, cex.axis = 1.5, 
               las = 1, ylab = "", xlab = "")

box(lwd = 2)

for(i in 2:18) {
  points(x = rnorm(dim(gr.raw.data.rank)[1], i, 0.05), 
         y = gr.raw.data.rank[,i], 
         pch = 21, bg = "lightgrey", col = "lightgrey", 
         lwd = 2, cex = 1.7)
}

mtext('Growth rate', side = 2, 
      outer = TRUE, cex = 2, line = -1.5, adj = 0.5)

# Major Axes

axis(side = 2, lwd.ticks = 2, cex.axis = 1.5, las = 1,at = c(0.0001, 0.001, 0.01, 0.1, 1), 
     labels = c(expression(paste("10"^"-4")), expression(paste("10"^"-3")), expression(paste("10"^"-2")),
     expression(paste("10"^"-1")), expression(paste("10"^" 0"))))
 
axis(side = 4, lwd.ticks = 2, cex.axis = 1.5, las = 1,
    at=c(0.0001, 0.001, 0.01, 0.1, 1), labels = F)
 

axis(side = 1, lwd.ticks = 2, cex.axis = 1.5, las = 1,
    labels = c(p.rank.name), at = c(1:18))


xtick<-p.rank.name
axis(side=1, at=xtick, labels = FALSE)
text(x=xtick,  par("usr")[3], 
     labels = xtick, srt = 45, pos = 1, xpd = TRUE)


# # Minor Axes
# axis(side = 2, tck = -0.015, labels = F, lwd.ticks = 2,
#        at = c(0.05, 0.1, 0.25, 0.5, 2))
# 
# axis(side = 4, tck = -0.015, labels = F, lwd.ticks = 2,
#        at = c(0.05, 0.1, 0.25, 0.5, 2))


# Adding means and 95 % confidence intervals
points(c(1:18), p.sort$p.means, pch = 21, col = "black", 
       bg = "NA", lwd = 2, cex = 2.5) 

arrows(x0 = c(1:18), y0 = p.sort$p.means, y1 = p.sort$p.95.LL, angle = 90,
       length = 0.1, lwd = 2)
arrows(x0 = c(1:18), y0 = p.sort$p.means, y1 = p.sort$p.95.UL, angle = 90,
       length=0.1, lwd = 2)

dev.off()
graphics.off() 

img <- readPNG("./figures/Ranked.P.png")
grid.raster(img)


```