title: "Phosphorus Traits"
author: "Jay T. Lennon, Mario Muscarella, Kali Bird"
date: "`r format(Sys.time(), '%d %B, %Y')`"
header-includes:
  - \usepackage{array}
output: pdf_document
geometry: margin=2.54cm
---

Generalist and specialist strategies of phosphorus acquistion by aquatic bacteria. In this file, the goal is to make a multi-panel heatmap representing three hypothetical scenarios and one observed outcome. The first three hypothetical scenarios come from resampling our data. Scenario one is a randomized data. Scenario two resamples our data assuming that resources are used similarly among strains. Scenario three resamples our data assuming the strains grow at different rates independent of phosphorus source. The last panel in the heatmap is our observed data.  

```{r}
rm(list=ls())
getwd()
setwd("~/GitHub/PhosTrait")
```

Load package for writing figures
```{r}
require("png")
require("vegan")
require("pvclust")
require("gplots")
require("gridGraphics")
require("grid")
require("gplots")
require("gridExtra")
```

Load data (note this is raw growth rates; not phylogenetically correct)
```{r}
# Raw growth rate data
gr.data <- read.csv("data/grraw.csv", sep=",", header=TRUE)
str(gr.data)
colnames(gr.data)[1]<-"isolate"

# Data standardizing - log10 tranformation
log.gr <- log10(gr.data[,2:19]+1)

# Data standardizing - divide by sum of species growth
gr.std <- gr.data[,2:19] / (apply(gr.data[,2:19], 1, sum))

# Names
resources <- colnames(gr.data[,2:19])
strains <- gr.data[,1]
```

Customize Color Palette
```{r}
jet.colors <- colorRampPalette(c("#00007F", "blue", "#007FFF", "cyan", 
                                 "#7FFF7F", "yellow", "#FF7F00", "red", 
                                 "#7F0000"))

gray.colors <- gray.colors(25, start = 0.9, end = 0.1, gamma = 1.5, alpha = 1)
```

Clustering parameters
```{r}
distance <- "euclidean" 
    # manhattan, correlation, uncentered,
    # abscor, euclidean, bray

# Identify agglomerative method
clust <- "average"   
    # complete = furthest neighbor; compact clusters, sensitive to outliers
    # average = UPGMA; considered robust
    # ward.D = popular, but makes clusters of equal size and sensitive to outliers
    # ward.D2 = dissimilarities are squared before clustering
    # mcquitty
    # median = downweights outliers
    # centroid = fairly robust
    # single = nearest neighbor; chaining problem
                    
```

Randomize growth rate data and make heatmap
```{r}
# Randomize growth rate data
log.gr.mat <- as.matrix(log.gr)
colnames(log.gr.mat) <- colnames(gr.data[,2:19])
rownames(log.gr.mat) <- gr.data[,1]
rand <- matrix(sample(log.gr.mat), nrow = nrow(log.gr))

# Make heatmap
rand.heat <- heatmap.2(as.matrix(rand), distfun = function(x) 
          vegdist(x, method = distance),
          hclustfun = function(x) hclust(x, method = clust), 
          col = gray.colors, trace = "none", density.info = "none", 
          labRow = strains, labCol = resources)
```

Calculate column-wise growth rate data (mean and sd) and make heatmap
```{r}
# Calculate column means
col.means <-colMeans(log.gr)

# Calculate standard deviation of columns
col.sd <- apply(log.gr, 2, sd)

# Create dataframe with mean column and sd
col.wise <- mapply(rnorm, nrow(log.gr), col.means, col.sd/30)
colnames(col.wise) <- resources

# Make heatmap
col.heat <- heatmap.2(as.matrix(col.wise), distfun = function(x) 
          vegdist(x, method = distance),
          hclustfun = function(x) hclust(x, method = clust), 
          col = gray.colors, trace = "none", density.info = "none",
          labRow = strains, labCol = resources)
```

Calculate row-wise growth rate data (mean and sd) and make heatmap
```{r}
# Calculate row means
row.means <-rowMeans(log.gr)

# Calculate standard deviation of rows
row.sd <- apply(log.gr, 1, sd)

# Create dataframe with mean rows and sd
row.wise <- t(mapply(rnorm,ncol(log.gr), row.means, row.sd/30))
colnames(row.wise) <- colnames(log.gr)
row.heat <- heatmap.2(as.matrix(row.wise), distfun = function(x) 
          vegdist(x, method = distance),
          hclustfun = function(x) hclust(x, method = clust), 
          col = gray.colors, trace = "none", density.info = "none",
          labRow = strains, labCol = resources)
```

Actual growth rate data as heatmap
```{r}
actual.heat <- heatmap.2(as.matrix(log.gr), distfun = function(x) 
          vegdist(x, method = distance), hclustfun = function(x) 
          hclust(x, method = clust), col = gray.colors, trace = "none", 
          density.info = "none", labRow = strains, 
          labCol = resources)
```

Make multipanel heatmap
```{r}
# Example: http://stackoverflow.com/questions/13081310/combining-multiple-complex-plots-as-panels-in-a-single-figure

# Make a list of dataframes for each scenario
heat.list <- list(rand, col.wise, row.wise, log.gr)

# Function for running through list
grab_grob <- function(){
  grid.echo()
  grid.grab()
}

gl <- lapply(1:4, function(i){
  heatmap.2(as.matrix(heat.list[[i]]), distfun = function(x) 
          vegdist(x, method = distance), hclustfun = function(x) 
          hclust(x, method = clust), col = gray.colors, trace = "none", 
          density.info = "none", labRow = strains, 
          labCol = resources, key = FALSE, keysize = 1.0, symkey = FALSE, 
          scale = "none", cexRow = 0.75, cexCol = 0.75,              
          lmat = rbind( c(0, 3), c(2,1), c(0,4) ), lhei = c(0.25, 4, 0.25 ),                 
  )
  grab_grob()
})

png(filename="~/GitHub/PhosTrait/figures/HeatMap.png", width = 2400, 
    height = 2400, res = 96*2)

grid.newpage()
grid.arrange(grobs = gl, ncol = 2, clip = TRUE)

dev.off()
graphics.off()
```
